\boldmath Disconnected pseudo-$C_\ell$ covariances for projected large-scale structure data

### 2019-06-24 draft/paper.tex
added plots in paper + new Fig for flat sky chi2 reldev mean & width

    \subsection{Quantitative comparison}\label{ssec:results.quant}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/run_sph_2b_NKA_TTTEEE_reldev_eigval_1stbin.pdf}
        \caption{Eigenvalues of the single-bin covariance matrices for all power spectra involving $\delta$ and $\gamma_E$. Results are shown for the sample covariance (solid blue), the NKA estimator (solid orange) and the spin-0 approximation (dashed green). The NKA and spin-0 estimators are able to recover the covariance eigenvalues with an accuracy $\lesssim5\%$.} \label{fig:eigv_1bin}
      \end{figure}
      \begin{figure}
        \centering
        \includegraphics[width=\textwidth]{./figures/run_sph_2b_1stbin_chi2_TT_TE_EE.pdf}
        \caption{$\chi^2$ distributions for the cases without $\gamma_B$-modes. We compare the distributions obtained with the covariance matrix computed directly from the simulated power spectra (blue) and the two different analytical methods: NKA (orange) and spin-0 (green). In addition, the theoretical $\chi^2$ distribution has been included (dashed red). The distribution extracted from the simulation follows the theoretical expectation almost perfectly in all cases. In the cases of $\delta\times\gamma_E$ and $\gamma_E\times\gamma_E$ we observe small shifts in the peak $\chi^2$, of $\sim2$-$4\%$, while they width is recovered accurately. We will show that these differences are due to the inaccuracy of the analytical approximations on the largest scales, and that they have a completely negligible impact on the final cosmological parameter constraints.}
        \label{fig:chi2_1bin} \end{figure}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/run_sph_2b_same_mask_Spin0_NKA_TTTEEE_Full_chi2.pdf}
        \caption{Same as Figure \ref{fig:chi2_1bin} for the combination of all correlations between $\delta$ and $\gamma_E$ for two redshift bins, where similar conclusions hold.}
        \label{fig:chi2_2bins}
      \end{figure}
      \begin{figure}
        \centering
        \includegraphics[width=\textwidth]{./figures/contaminants_diag_cov.pdf}
        \caption{Diagonal of the covariance matrix estimated from simulations without contaminants (solid blue), with 200 contaminants deprojected (dashed orange) and the NKA estimator (solid green). The differences between the simulated cases are significantly smaller than those associated with the NKA estimator. The NKA method is therefore equally valid to approximate the covariance even in the presence of contaminant deprojection.}
        \label{fig:diag_conts}
      \end{figure}
      In order to quantify the validity of our analytical approximations, we need to compare the NKA covariance with the sample covariance estimated from the simulations. However, comparing two matrices is not as straightforward as comparing their matrix elements one by one. The covariance between far-away bandpowers is expected to be very close to zero, and therefore a direct comparison of those elements would easily yield large relative differences simply due to the statistical noise in the sample covariance matrix. We will therefore quantify the differences between the different covariances making use of scalar quantities formed from them. The impact of the analytical approximations on the final parameter constraints will then be described in detail in Section \ref{ssec:results.cosmo}.

@@

      Another scalar quantity that can be used to compare different covariances is the $\chi^2$. For a given random data vector ${\bf d}$ with mean ${\bf m}$ and covariance matrix ${\rm Cov}$, this is given by
      \begin{equation}
        \chi^2=\left({\bf d}-{\bf m}\right)^T\cdot{\rm Cov}^{-1}\cdot\left({\bf d}-{\bf m}\right).
      \end{equation}
      We compute this quantity for a data vector ${\bf d}$ composed of different auto- and cross-correlations for each of the 20,000 Gaussian simulations, with ${\bf m}$ given by the mean over all simulations and different choices of covariance matrix. Figure \ref{fig:chi2_1bin} shows the distribution of $\chi^2$ values for the three non-zero power spectra in the case of a single bin: $\delta$-$\delta$, $\delta$-$\gamma_E$ and $\gamma_E$-$\gamma_E$. The histograms show the distribution for the sample covariance matrix (blue), and the analytical NKA and Spin-0 estimators (orange and green respectively). We additionally plot the theoretical $\chi^2$ distribution under the assumption that the underlying data vector is Gaussianly distributed (red dashed lines). In the simplest case of purely spin-0 quantities (leftmost panel), we find an excellent agreement between the different distributions. In the cases involving the spin-2 fields, we see noticeable differences between the distributions found with the sample covariance and the approximate ones.
      These differences are small, corresponding to less of a $2$ and $4\%$ shift in the mean $\chi^2$ for the $\delta$-$\gamma_E$ and $\gamma_E$-$\gamma_E$ cases respectively, and a negligible variation in the width of the distributions. We therefore expect these differents to have a negligible effect on the posterior parameter distributions, as we show explicitly in Section \ref{ssec:results.cosmo}. The fact that these differences appear only for power spectra involving spin-2 quantities indicate that the NKA and spin-0 methods are imperfect at describing the additional mode coupling caused in the presence of a mask for higher-spin fields. This is rather obvious in the case of the spin-0 approximation but, interestingly, we find that the NKA and Spin-0 predictions yield results that are almost indistinguishable from each other. We therefore conclude that the additional approximation made in the NKA method for spin-2 fields -- neglecting the spatial derivatives of the mask -- is effectively equivalent to ignoring the spin nature of the fields involved. For completeness, Figure \ref{fig:chi2_2bins} shows the distribution of $\chi^2$ values for a data vector composed of all possible auto- and cross-correlations of $\delta$ and $\gamma_E$ for the case of two redshift bins, where similar conclusions hold.

@@

    \begin{figure}
      \centering
      \includegraphics[width=\textwidth]{./figures/run_chi2_TT_TE_EE_TB_EB_BB.pdf}
      \caption{Same as Figure \ref{fig:chi2_1bin} for flat-sky fields. This figure additionally shows the $\chi^2$ distributions for power spectra involving $B$-modes. For the smaller scales covered by these flat-sky simulations, the NKA and spin-0 estimators work remarkably well, even for $B$-mode spectra.}
      \label{fig:chi2_1bin_flat}
    \end{figure}
    We have repeated the analysis described in Section \ref{sec:results} on flat-sky realizations, making use of the flat-sky implementation of {\tt NaMaster}. We generate Gaussian realizations of the galaxy overdensity and shear maps making use of flat-sky extensions of the methods described in Section \ref{ssec:results.sims}. In this case we use a high-resolution mask constructed from the bright-object mask distributed with the first data release of the HSC collaboration \cite{2018PASJ...70S...4A} for the VVDS field, which is shown in Fig. \ref{fig:mask_flat}.

    We find similar levels of accuracy in the NKA and spin-0 estimators compared to the curved-sky case. The higher resolution and smaller area of these simulations allow us to focus on the small-scale galaxy clustering and lensing power spectra, covering the range of multipoles $\ell\in(120,17640)$ in constant bandpowers of width $\Delta \ell=240$. On these small scales, the shear power spectrum is more dominated by noise (e.g. see Fig. \ref{fig:cl-2bins}), and therefore there is roughly the same power in $E$ and $B$ modes. This reduces the sensitivity of the method to an inaccurate treatment of $E/B$ leakage, and the agreement between the sample covariance matrix and the NKA estimator improves significantly. This can be seen in Fig. \ref{fig:chi2_1bin_flat}, which shows the $\chi^2$ distributions for all possible power spectra (including those including $B$ modes) in the case of a single redshift bin. In all cases we find a good agreement between the $\chi^2$ distributions derived from all covariance matrix estimates (sample covariance, NKA and spin-0 in blue, orange and green respectively), which also accurately follow the expected $\chi^2$ distribution for the corresponding number of degrees of freedom (red dashed line).

  \section{Software implementation}\label{app:namaster}
    In addition to the code functionality described in Section 3 of \cite{2019MNRAS.484.4127A}, we have now included the capability to estimate Gaussian covariance matrices using the NKA method. This functionality is structured around a {\tt python} class called {\tt NmtCovarianceWorkspace}. These objects are used to compute and store the covariance mode-coupling coefficients in Eqs. \ref{eq:coeff_cmcm} and \ref{eq:coeff_cmcm_flat}. They are initialized from two pairs of fields corresponding to the two power spectra for which the covariance is required. Once initialized, these coefficients can be reused for any other set of fields with the same combination of sky masks. {\tt NaMaster} then provides routines to estimate covariance matrix elements making use of the coupling coefficients stored in a {\tt NmtCovarianceWorkspace} object and best-guess power spectra for the fields involved using Eqs. \ref{eq:cov_sph} and \ref{eq:cov_flat}. Further details about the implementation and practical examples can be found in \url{https://namaster.readthedocs.io/en/latest/sample_covariance.html}.

>>>>

    \subsection{Quantitative comparison}\label{ssec:results.quant}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/run_sph_2b_NKA_TTTEEE_reldev_eigval_1stbin.pdf}
        \caption{Eigenvalues of the single-bin covariance matrices for all power spectra involving $\delta$ and $\gamma_E$. Results are shown for the sample covariance (solid blue), the NKA estimator (solid orange) and the spin-0 approximation (dashed green). The NKA and spin-0 estimators are able to recover the covariance eigenvalues with an accuracy $\lesssim5\%$.} \label{fig:eigv_1bin}
      \end{figure}
      \begin{figure}
        \centering
        \includegraphics[width=\textwidth]{./figures/run_sph_2b_1stbin_chi2_TT_TE_EE.pdf}
        \caption{$\chi^2$ distributions for the cases without $\gamma_B$-modes. We compare the distributions obtained with the covariance matrix computed directly from the simulated power spectra (blue) and the two different analytical methods: NKA (orange) and spin-0 (green). In addition, the theoretical $\chi^2$ distribution has been included (dashed red). The distribution extracted from the simulation follows the theoretical expectation almost perfectly in all cases. In the cases of $\delta\times\gamma_E$ and $\gamma_E\times\gamma_E$ we observe small shifts in the peak $\chi^2$, of $\sim2$-$4\%$, while they width is recovered accurately. We will show that these differences are due to the inaccuracy of the analytical approximations on the largest scales, and that they have a completely negligible impact on the final cosmological parameter constraints.}
        \label{fig:chi2_1bin} \end{figure}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/run_sph_2b_1stbin_chi2_mean_cov_diff.pdf}
        \caption{$\chi^2$ distributions' mean and width relative difference
          between that obtained with a covariance matrix estimated from
          analytical methods (NKA and Spin-0) and simulations. This is for the
          case with 1 bin.}
        \label{fig:chi2_1bin_diffs}
      \end{figure}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/run_sph_2b_Spin0_NKA_TTTEEE_Full_chi2.pdf}
        \caption{Same as Figure \ref{fig:chi2_1bin} for the combination of all correlations between $\delta$ and $\gamma_E$ for two redshift bins, where similar conclusions hold.}
        \label{fig:chi2_2bins}
      \end{figure}
      \begin{table}
        \centering
        \begin{tabular}{|l|c|c|}
          \hline
              & NKA & Spin-0 \\
          \hline
          Rel. Dev. Mean [\%]  & -2.8 & -2.8 \\ 
          Rel. Dev. Width [\%] & 6.5  & 6.9  \\
          \hline
        \end{tabular}
        \caption{$\chi^2$ mean and width relative difference between that
          obtained with a covariance matrix estimated from simulations and
          the analytical methods NKA and Spin-0.} 
        \label{tab:chi2_2bins_diffs}
      \end{table}
      \begin{figure}
        \centering
        \includegraphics[width=\textwidth]{./figures/contaminants_diag_cov.pdf}
        \caption{Diagonal of the covariance matrix estimated from simulations without contaminants (solid blue), with 200 contaminants deprojected (dashed orange) and the NKA estimator (solid green). The differences between the simulated cases are significantly smaller than those associated with the NKA estimator. The NKA method is therefore equally valid to approximate the covariance even in the presence of contaminant deprojection.}
        \label{fig:diag_conts}
      \end{figure}
      In order to quantify the validity of our analytical approximations, we need to compare the NKA covariance with the sample covariance estimated from the simulations. However, comparing two matrices is not as straightforward as comparing their matrix elements one by one. The covariance between far-away bandpowers is expected to be very close to zero, and therefore a direct comparison of those elements would easily yield large relative differences simply due to the statistical noise in the sample covariance matrix. We will therefore quantify the differences between the different covariances making use of scalar quantities formed from them. The impact of the analytical approximations on the final parameter constraints will then be described in detail in Section \ref{ssec:results.cosmo}.

@@

      Another scalar quantity that can be used to compare different covariances is the $\chi^2$. For a given random data vector ${\bf d}$ with mean ${\bf m}$ and covariance matrix ${\rm Cov}$, this is given by
      \begin{equation}
        \chi^2=\left({\bf d}-{\bf m}\right)^T\cdot{\rm Cov}^{-1}\cdot\left({\bf d}-{\bf m}\right).
      \end{equation}
      We compute this quantity for a data vector ${\bf d}$ composed of different auto- and cross-correlations for each of the 20,000 Gaussian simulations, with ${\bf m}$ given by the mean over all simulations and different choices of covariance matrix. Figure \ref{fig:chi2_1bin} shows the distribution of $\chi^2$ values for the three non-zero power spectra in the case of a single bin: $\delta$-$\delta$, $\delta$-$\gamma_E$ and $\gamma_E$-$\gamma_E$. The histograms show the distribution for the sample covariance matrix (blue), and the analytical NKA and Spin-0 estimators (orange and green respectively). We additionally plot the theoretical $\chi^2$ distribution under the assumption that the underlying data vector is Gaussianly distributed (red dashed lines). In the simplest case of purely spin-0 quantities (leftmost panel), we find an excellent agreement between the different distributions. In the cases involving the spin-2 fields, we see noticeable differences between the distributions found with the sample covariance and the approximate ones.
      These differences are small, corresponding to less of a $2$ and $4\%$
      shift in the mean $\chi^2$ for the $\delta$-$\gamma_E$ and
      $\gamma_E$-$\gamma_E$ cases respectively, and $\lesssim 2\%$ variation
      in the width of the distributions (see
      Figure~\ref{fig:chi2_1bin_diffs}).  We therefore expect these
      differences to have a negligible effect on the posterior parameter
      distributions, as we show explicitly in Section
      \ref{ssec:results.cosmo}. The fact that these differences appear only
      for power spectra involving spin-2 quantities indicate that the NKA and
      spin-0 methods are imperfect at describing the additional mode coupling
      caused in the presence of a mask for higher-spin fields. This is rather
      obvious in the case of the spin-0 approximation but, interestingly, we
      find that the NKA and Spin-0 predictions yield results that are almost
      indistinguishable from each other. We therefore conclude that the
      additional approximation made in the NKA method for spin-2 fields --
      neglecting the spatial derivatives of the mask -- is effectively
      equivalent to ignoring the spin nature of the fields involved. For
      completeness, Figure \ref{fig:chi2_2bins} shows the distribution of
      $\chi^2$ values for a data vector composed of all possible auto- and
      cross-correlations of $\delta$ and $\gamma_E$ for the case of two
      redshift bins, where similar conclusions hold.

@@

    \begin{figure}
      \centering
      \includegraphics[width=\textwidth]{./figures/run_chi2_TT_TE_EE_TB_EB_BB.pdf}
      \caption{Same as Figure \ref{fig:chi2_1bin} for flat-sky fields. This figure additionally shows the $\chi^2$ distributions for power spectra involving $B$-modes. For the smaller scales covered by these flat-sky simulations, the NKA and spin-0 estimators work remarkably well, even for $B$-mode spectra.}
      \label{fig:chi2_1bin_flat}
    \end{figure}
      \begin{figure}
        \centering
        \includegraphics[width=0.6\textwidth]{./figures/run_chi2_mean_cov_diff.pdf}
        \caption{$\chi^2$ distributions' mean and width relative difference
          between that obtained with a covariance matrix estimated from
          analytical methods (NKA and Spin-0) and simulations. This is for the
          case with 1 bin in the flat-sky approximation.}
        \label{fig:chi2_1bin_flat_diffs}
      \end{figure}
    We have repeated the analysis described in Section \ref{sec:results} on flat-sky realizations, making use of the flat-sky implementation of {\tt NaMaster}. We generate Gaussian realizations of the galaxy overdensity and shear maps making use of flat-sky extensions of the methods described in Section \ref{ssec:results.sims}. In this case we use a high-resolution mask constructed from the bright-object mask distributed with the first data release of the HSC collaboration \cite{2018PASJ...70S...4A} for the VVDS field, which is shown in Fig. \ref{fig:mask_flat}.

    We find similar levels of accuracy in the NKA and spin-0 estimators
    compared to the curved-sky case. The higher resolution and smaller area of
    these simulations allow us to focus on the small-scale galaxy clustering
    and lensing power spectra, covering the range of multipoles
    $\ell\in(120,17640)$ in constant bandpowers of width $\Delta \ell=240$. On
    these small scales, the shear power spectrum is more dominated by noise
    (e.g. see Fig. \ref{fig:cl-2bins}), and therefore there is roughly the
    same power in $E$ and $B$ modes. This reduces the sensitivity of the
    method to an inaccurate treatment of $E/B$ leakage, and the agreement
    between the sample covariance matrix and the NKA estimator improves
    significantly. This can be seen in Figs.~\ref{fig:chi2_1bin_flat} and
    \ref{fig:chi2_1bin_flat_diffs}, which
    shows the $\chi^2$ distributions for all possible power spectra (including
    those including $B$ modes) in the case of a single redshift bin. In all
    cases we find a good agreement (difference $\lesssim 2 \%$) between the
    $\chi^2$ distributions derived from all covariance matrix estimates
    (sample covariance, NKA and spin-0 in blue, orange and green
    respectively), which also accurately follow the expected $\chi^2$
    distribution for the corresponding number of degrees of freedom (red
    dashed line).     
  \section{Software implementation}\label{app:namaster}
    In addition to the code functionality described in Section 3 of \cite{2019MNRAS.484.4127A}, we have now included the capability to estimate Gaussian covariance matrices using the NKA method. This functionality is structured around a {\tt python} class called {\tt NmtCovarianceWorkspace}. These objects are used to compute and store the covariance mode-coupling coefficients in Eqs. \ref{eq:coeff_cmcm} and \ref{eq:coeff_cmcm_flat}. They are initialized from two pairs of fields corresponding to the two power spectra for which the covariance is required. Once initialized, these coefficients can be reused for any other set of fields with the same combination of sky masks. {\tt NaMaster} then provides routines to estimate covariance matrix elements making use of the coupling coefficients stored in a {\tt NmtCovarianceWorkspace} object and best-guess power spectra for the fields involved using Eqs. \ref{eq:cov_sph} and \ref{eq:cov_flat}. Further details about the implementation and practical examples can be found in \url{https://namaster.readthedocs.io/en/latest/sample_covariance.html}.
